Bandit Multiclass Linear Classification: \\ Efficient Algorithms for the Separable Case

### 2019-01-23 ICML-submission/mistake-lower-bound-for-ignorant-algorithms.tex
Clean-up Appendix I

Under the assumption that the examples lie in in the unit ball of $\R^d$ and are
weakly linearly separable with margin $\gamma$, we show that any ignorant
algorithm must make $\Theta\left(\min\{\sqrt{T},
(1/\gamma)^{\Theta(d)}\}\right)$ mistakes in the worst case. In other words, an
algorithm that achieves a better mistake bound cannot ignore examples on which
it makes a mistake and it must make a meaningful update on such examples.

The ignorant algorithm is formally defined as follows.
\begin{definition}
\label{definition:ignorant-algorithm}
An algorithm $\calA$ for \textsc{Online Multiclass Linear Classification with
Bandit Feedback} is called \emph{ignorant} if for all $t \in \{1,2,\dots,T\}$,
$\widehat y_t$ is drawn from some distribution that is jointly determined by
\begin{enumerate}
\item $x_t$,
\item $H_t \triangleq \left( (x_{\tau_1}, y_{\tau_1}), (x_{\tau_2}, y_{\tau_2}), \ldots, (x_{\tau_{n_t}}, y_{\tau_{n_t}}) \right)$,
where $1 \le \tau_1 < \dots < \tau_{n_t} < t$, and $\{\tau_1, \ldots, \tau_{n_t}\}$
are the rounds in which the algorithm predicts correctly (that is,
$\{\tau_1, \dots, \tau_{n_t}\} = \{1 \le \tau < t ~:~ \widehat y_\tau=y_\tau\}$).
\end{enumerate}
We use $p_t(\cdot|x_t)\in \{p\in \R_+^K ~:~ p^\top\mathbf{1}=1\}$ to denote
the distribution $\widehat y_t$ is drawn from. Note that we let the current
feature vector $x_t$ be an argument to $p_t$ to incorporate its dependency on
$x_t$. Therefore, for an ignorant algorithm, the function $p_t(\cdot|\cdot)$ is
only determined by $H_t$.
\end{definition}

An ignorant algorithm does not learn from mistakes in the following sense.
Suppose that an ignorant algorithm predicts incorrectly at time $t$. Then at
time $t+1$, if it is given the same feature vector $x_{t+1}=x_t$, it will
predict the label with the same distribution
$p_{t+1}(\cdot|x_{t+1})=p_{t+1}(\cdot|x_t)=p_t(\cdot|x_t)$.

% Below we describe
%the difficulty of learning from mistakes using a class of natural approaches
%under the weak separability assumption. Consider an approach similar to OBAMA:
%in each round $t$, we create a \textit{convex} loss function $\widehat \ell_t:
%\R^{K\times d} \to \R$ that gives a loss value $\widehat
%\ell_t(W)$ to all linear models $W \in \R^{K\times d}$ ($W=[w_1 \dots
%w_K]^\top$ for $w_1, w_2, \dots, w_K$ defined in
%Definition~\ref{definition:linear-separability}). The requirement for
%$\widehat{\ell}_t$ to be convex is due to computational efficiency. To let
%$\widehat{\ell}_t$ reflect the loss, if $\widehat y_t \neq y_t$, we want to
%assign larger loss values $\widehat{\ell_t}(W)$ to the inconsistent models
%$\mathcal{W}=\{W: \ip{w_{\widehat y_t}}{x_t} \ge \ip{w_i}{x_t}, \forall
%i\in[K]\}$, while for other $W$'s, we want $\widehat{\ell_t}(W)$ to be smaller.
%The following fact can be verified when $K\ge 3$, which suggests there is no
%way to define the loss $\widehat \ell_t$ that satisfies the above properties
%constraints: there exists $W_1, W_2, W_3$, where $W_2$ is in the line segment
%between $W_1$ and $W_3$, such that $W_2\in \mathcal{W}$, while $W_1, W_3 \notin
%\mathcal{W}$. On one hand, by convexity, we should have
%$\widehat{\ell}_t(W_2)\le \max\{\widehat{\ell}_t(W_1),
%\widehat{\ell}_t(W_3)\}$; on the other hand, to penalize $\mathcal{W}$, we
%should have $\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_1)$ and
%$\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_3)$. Clearly, these two requirements
%contradict each other.

The following theorem states the mistake lower bound for ignorant algorithms.

\begin{theorem}[Mistake lower bound for ignorant algorithms]
\label{theorem:ignorant-lower-bound}
Suppose $\calA$ is an ignorant algorithm for \textsc{Online Multiclass Linear
Classification with Bandit Feedback}. There exists an adversary that can
sequentially choose labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)
\in \R^d\times \{1,2\}$ so that the examples are $\gamma$-strongly linearly
separable, $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le 1$, and the expected
number of mistakes made by $\calA$ is at least
$$
\frac{1}{10}\min\left\{\sqrt{T}, \left(\frac{1}{160\gamma}\right)^{\frac{d-2}{4}} \right\} \; .
$$
\end{theorem}

@@

\begin{proof}[Proof for Theorem~\ref{theorem:ignorant-lower-bound} (Part I: Mistake lower bound)]
We consider the strategy for the adversary described in Algorithm~\ref{algorithm:adversary-strategy}.
\begin{algorithm}[h]
\caption{\textsc{Adversary's strategy}}
\label{algorithm:adversary-strategy}
\begin{algorithmic}[1]
{
\STATE{\textbf{Define} $q_0=\max\left\{\frac{1}{\sqrt{T}}, \frac{1}{\sqrt{N}}\right\}$, where $N$ is defined in Lemma~\ref{lemma:embed_d_gamma}. Also, let $v_1, \ldots, v_N$ be those defined in Lemma~\ref{lemma:embed_d_gamma}.}
\STATE{\textbf{Initialize} $\textsc{phase}= 1$. }
\FOR{$t=1,\dots,\min\{T,N\}$}
    \IF{$\textsc{phase}=1$}
       \IF{$p_t(1|v_t)\ge 1-q_0$}
          \STATE{$(x_t, y_t)\leftarrow (v_t, 1)$}
       \ELSE
          \STATE{$(x_t, y_t)\leftarrow (v_t, 2)$}
          \STATE $\textsc{phase}\leftarrow 2$
       \ENDIF
    \ELSIF{$\textsc{phase}=2$}
       \STATE $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$.
    \ENDIF
\ENDFOR
\FOR{$t=\min\{T,N\}+1, \ldots, T$}
    \STATE $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$.
\ENDFOR

}
\end{algorithmic}
\end{algorithm}
For this strategy, we prove the mistake lower bound of any ignorant algorithm.
Define the indicators
\begin{align*}
A_t &= \indicator{\forall \tau\le t, p_\tau(1|x_\tau)<1-q_0} \\
B_t &= \indicator{\exists \tau\le t, p_\tau(1|x_\tau)\ge 1-q_0
 \text{\ and\ } \forall s\in[\tau,t), \widehat y_s \neq y_s}.
\end{align*}
Then we have
\begin{align}
& \mathbf{E}\left[\sum_{t=1}^{T} \indicator{\widehat y_t\neq y_t}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \indicator{\widehat y_t\neq y_t}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \indicator{\widehat y_t\neq y_t}A_t\right] \nonumber \\
& \qquad + \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \indicator{\widehat y_t\neq y_t}B_t\right] \nonumber \\
& = \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \mathbf{E}_t\left[\indicator{\widehat y_t\neq y_t}A_t\right]\right] + \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} B_{t+1}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} q_0 A_t\right] + \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} B_{t+1}\right],
\label{equation:mistake_lower_bound_temp}
\end{align}
where in the first inequality we use the fact that $A_t$ and $B_t$ cannot be $1$ simultaneously; in the equality, we use
$B_{t+1}=B_t\indicator{\widehat y_t\neq y_t}$ by $B_t$'s definition; in the last
inequality we use the fact that when $t\le N$ and $A_t=1$, it must be $y_t=1$ and $\Pr[\widehat y_t=y_t]=p_t(1|x_t)<1-q_0$.
Now, denote $T_1 = \argmin_{\tau} \{ p_\tau(1|x_\tau) \ge 1-q_0 \}$ (if such
$\tau$ does not exist or is larger than $\min\{T,N\}$, we simply let
$T_1=\min\{T,N\}+1$). Then $A_t=1$ for all $t\le T_1-1$.

@@

\begin{align*}
& \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}}B_{t+1}\right] \\
& \ge \mathbf{E}\left[\sum_{t=T_1}^{\min\{ T,N \}} (1-q_0)^{t-T_1+1} \right] \\
& = \frac{1-q_0}{q_0}\mathbf{E}\left[1-(1-q_0)^{\min\{T,N\}-T_1+1}\right] \; .
\end{align*}
Combining \eqref{equation:mistake_lower_bound_temp} and the above inequalities, we get
\begin{multline}
\label{equation:mistake_lower_bound_temp2}
\mathbf{E} \left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}\right] \ge
\\ \mathbf{E}\left[q_0(T_1-1)+\frac{1-q_0}{q_0}\left(1-(1-q_0)^{\min\{T,N\}-T_1+1}\right)\right].
\end{multline}
In the case $T_1\ge \frac{1}{2}\min\{T,N\}+1$, the right-hand side of
\eqref{equation:mistake_lower_bound_temp2} is lower bounded by
$\frac{1}{2}q_0\min\{T,N\}=\frac{1}{2}\min\{\sqrt{T}, \sqrt{N}\}$; in the case
$T_1< \frac{1}{2}\min\{T,N\}+1$, it is lower bounded by
\begin{align*}
& \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2}\min\{T,N\}}\right) \\
& = \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2q_0^2}}\right) \\
& \ge \frac{1-\frac{1}{\sqrt{2}}}{q_0}\left(1-\frac{1}{\sqrt{e}}\right) \\
& \ge \frac{1}{10}\min\{\sqrt{T}, \sqrt{N}\} \; .
\end{align*}
Combining both cases concludes the proof.
\end{proof}

\begin{proof}[Proof for Theorem~\ref{theorem:ignorant-lower-bound} (Part II: linear separability)]
Observe that in Algorithm~\ref{algorithm:adversary-strategy}, in the beginning
the adversary always set the label to $1$; once it sets $y_t=2$ at some $t$, it
lets $(x_s, y_s)=(x_t, y_t)=(v_t, 2)$ for all $s\ge t$. Consider $w_2=u_t$ and
$w_1=-u_t$, for $u_t$ as defined in Lemma~\ref{lemma:embed_d_gamma}. Then by
Lemma~\ref{lemma:embed_d_gamma}, we have $\ip{w_2}{x_s} = \ip{u_t}{v_t} \ge
\gamma, \forall s \ge t$ and $\ip{w_2}{x_s} = \ip{u_t}{v_s} \le - \gamma,
\forall s< t$; also $\ip{w_1}{x_s} \le -\gamma, \forall s \ge t$ and
$\ip{w_1}{x_s} \ge  \gamma, \forall s<t$. Thus, the example set is always
$\gamma$-strongly separable.
\end{proof}

>>>>

Under the assumption that the examples lie in in the unit ball of $\R^d$ and are
weakly linearly separable with margin $\gamma$, we show that any ignorant
algorithm must make at least $\Omega \left( \left(\frac{1}{160 \gamma}\right)^{(d-2)/4} \right)$
mistakes in the worst case. In other words, an algorithm that achieves a better
mistake bound cannot ignore examples on which it makes a mistake and it must
make a meaningful update on such examples.

To formally define ignorant algorithms, we define the conditional distribution
from which an algorithm draws its predictions. Formally, given an algorithm
$\calA$ and an adversarial strategy, we define
\begin{multline*}
p_t(y|x) = \\
\Pr[y_t = y ~|~ (x_1, y_1), (x_2, y_2) \dots, (x_{t-1}, y_{t-1}), x_t = x] \; .
\end{multline*}
In other words, in any round $t$, conditioned on the past $t-1$ rounds, the
algorithm $\calA$ chooses $y_t$ from probability distribution $p_t(\cdot|x_t)$.
Formally, $p_t$ is a function $p:\{1,2,\dots,K\} \times \R^d \to [0,1]$
such that $\sum_{y=1}^K p_t(y|x) = 1$ for any $x \in \R^d$.

\begin{definition}[Ignorant algorithm]
An algorithm $\calA$ for \textsc{Online Multiclass Linear Classification with
Bandit Feedback} is called \emph{ignorant} if for every $t=1,2,\dots,T$,
$p_t$ is determined solely by the sequence
$(x_{a_1}, y_{a_1}), (x_{a_2}, y_{a_2}), \dots, (x_{a_n}, y_{a_n})$
in the rounds $1 \le a_1 < a_2 < \dots < a_n < t$ in which
the algorithm makes a correct prediction.
\end{definition}

An equivalent definition of an ignorant algorithm is that the memory state of
the algorithm does not change after it makes a mistake. Equivalently,
the memory state of an ignorant algorithm is completely determined
by the sequence of labeled examples on which it made correct prediction.

To explain the definition, consider an ignorant algorithm $\calA$. Suppose that
on a sequence of examples $(x_1, y_1), (x_2, y_2), \dots, (x_{t-1}, y_{t-1})$
generated by some adversary the algorthm $\calA$ makes correct predictions in
rounds $a_1, a_2, \dots, a_n$ where $1 \le a_1 < a_2 < \dots < a_n < t$ and
errors on rounds $\{1,2,\dots,t-1\} \setminus \{a_1, a_2, \dots, a_n\}$. Suppose
that on another sequence of examples $(x_1', y_1'), (x_2', y_2'), \dots,
(x_{s-1}', y_{s-1}')$ generated by another adversary the algorithm $\calA$ makes
correct predictions in rounds $b_1, b_2, \dots, b_n$ where $1 \le b_1 < b_2 <
\dots < b_n < s$ and errors on rounds $\{1,2,\dots,s-1\} \setminus \{b_1, b_2,
\dots, b_n\}$. Futhermore, suppose
\begin{align*}
(x_{a_1}, y_{a_1}) &= (x'_{b_1}, y'_{b_1}) \; , \\
(x_{a_2}, y_{a_2}) &= (x'_{b_2}, y'_{b_2}) \; , \\
\vdots \\
(x_{a_n}, y_{a_n}) &= (x'_{b_2}, y'_{b_n}) \; .
\end{align*}
Then, the definition
\begin{multline*}
\Pr[y_t = y ~|~ (x_1, y_1), (x_2, y_2) \dots, (x_{t-1}, y_{t-1}), x_t = x] = \\
\Pr[y_t' = y ~|~ (x_1', y_1'), (x_2', y_2') \dots, (x_{t-1}', y_{t-1}'), x_t' = x] \; .
\end{multline*}
Note that the sequences $(x_1, y_1)$, $(x_2, y_2)$, $\dots$, $(x_{t-1},
y_{t-1})$ and $(x_1', y_1')$, $(x_2', y_2')$, $\dots$, $(x_{s-1}', y_{s-1}')$
might have different lengths and and $\calA$ might error in different sets of
rounds. As a special case, if an ignorant algorithm makes a mistake in round $t$
then $p_{t+1}=p_t$.

Our main result is the following lower bound on the expected number of mistakes
for ignorant algorithms.

\begin{theorem}[Mistake lower bound for ignorant algorithms]
\label{theorem:ignorant-lower-bound}
Let $\gamma \in (0,1)$ and let $d$ be a positive integer. Suppose $\calA$ is an
ignorant algorithm for \textsc{Online Multiclass Linear Classification with
Bandit Feedback}. There exists $T$ and an adversary that sequentially chooses
labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in \R^d\times
\{1,2\}$ such that the examples are strongly linearly separable with magin
$\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le 1$, and the expected
number of mistakes made by $\calA$ is at least
$$
\frac{1}{10} \left(\frac{1}{160\gamma}\right)^{\frac{d-2}{4}} \; .
$$
\end{theorem}

@@

\begin{proof}
We consider the strategy for the adversary described in
Algorithm~\ref{algorithm:adversary-strategy}.

\begin{algorithm}
\caption{\textsc{Adversary's strategy}}
\label{algorithm:adversary-strategy}
\textbf{Define} $T=N$ and $v_1, v_2, \dots, v_N$ as in Lemma~\ref{lemma:embed_d_gamma}.\\
\textbf{Define} $q_0=\frac{1}{\sqrt{T}}$. \\
\textbf{Initialize} $\textsc{phase}= 1$. \\
\For{$t=1,2,\dots,T$}{
    \If{$\textsc{phase}=1$}{
       \If{$p_t(1|v_t)\ge 1-q_0$}{
          $(x_t, y_t)\leftarrow (v_t, 1)$
        }
       \Else{
          $(x_t, y_t)\leftarrow (v_t, 2)$ \\
          $\textsc{phase}\leftarrow 2$
       }
    }
    \Else{
         $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$
    }
}
\end{algorithm}

Define the indicators
\begin{align*}
A_t &= \indicator{\forall \tau\le t, p_\tau(1|x_\tau)<1-q_0} \\
B_t &= \indicator{\exists \tau\le t, p_\tau(1|x_\tau)\ge 1-q_0
 \text{\ and\ } \forall s\in[\tau,t), \widehat y_s \neq y_s}.
\end{align*}
Then we have
\begin{align}
& \mathbf{E}\left[\sum_{t=1}^{T} \indicator{\widehat y_t\neq y_t}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}A_t\right] \nonumber \\
& \qquad + \mathbf{E}\left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}B_t\right] \nonumber \\
& = \mathbf{E}\left[\sum_{t=1}^T \mathbf{E}_t\left[\indicator{\widehat y_t\neq y_t}A_t\right]\right] + \mathbf{E}\left[\sum_{t=1}^T B_{t+1}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^T q_0 A_t\right] + \mathbf{E}\left[\sum_{t=1}^T B_{t+1}\right],
\label{equation:mistake_lower_bound_temp}
\end{align}
where in the first inequality we use the fact that $A_t$ and $B_t$ cannot be $1$
simultaneously; in the equality, we use $B_{t+1}=B_t\indicator{\widehat y_t\neq
y_t}$ by $B_t$'s definition; in the last inequality we use the fact that when
$t\le N$ and $A_t=1$, it must be $y_t=1$ and $\Pr[\widehat
y_t=y_t]=p_t(1|x_t)<1-q_0$. Now, denote $T_1 = \argmin_{\tau} \{
p_\tau(1|x_\tau) \ge 1-q_0 \}$ (if such $\tau$ does not exist or is larger than
$T$, we simply let $T_1=T+1$). Then $A_t=1$ for all $t\le
T_1-1$.

@@

\begin{align*}
\mathbf{E}\left[\sum_{t=1}^T B_{t+1}\right]
& \ge \mathbf{E}\left[\sum_{t=T_1}^T (1-q_0)^{t-T_1+1} \right] \\
& = \frac{1-q_0}{q_0}\mathbf{E}\left[1-(1-q_0)^{T - T_1 + 1}\right] \; .
\end{align*}
Combining \eqref{equation:mistake_lower_bound_temp} and the above inequalities, we get
\begin{multline}
\label{equation:mistake_lower_bound_temp2}
\mathbf{E} \left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}\right] \ge
\\ \mathbf{E}\left[q_0(T_1-1)+\frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-T_1+1}\right)\right].
\end{multline}
In the case $T_1\ge \frac{1}{2}T + 1$, the right-hand side of
\eqref{equation:mistake_lower_bound_temp2} is lower bounded by
$\frac{1}{2}q_0 T = \frac{1}{2} \sqrt{T}$. In the case
$T_1< \frac{1}{2}T+1$, it is lower bounded by
\begin{align*}
& \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2}T}\right) \\
& = \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2q_0^2}}\right) \\
& \ge \frac{1-\frac{1}{\sqrt{2}}}{q_0}\left(1-\frac{1}{\sqrt{e}}\right) \\
& \ge \frac{1}{10} \sqrt{T} \; .
\end{align*}

Observe that in phase 1, the labels are equal to $1$ and in phase 2 the labels
are equal to $2$. Let $t$ be the round in which the switch happens. (If it never
happens, define $t=T+1$.) Note that $(x_t, y_t)=(x_{t+1}, y_{t+1})= \dots =
(x_T, y_T) = (v_t, 2)$. Consider the vectors $u_1, u_2, \dots, u_N$ as defined
in Lemma~\ref{lemma:embed_d_gamma} We claim that $w_1=-u_t/2$ and $w_2=u_t/2$
satisfy the conditions of strong linear separabaility.

Clearly $\norm{w_1}^2 + \norm{w_2}^2 \le (\norm{w_1} + \norm{w_2})^2 \le
(\frac{1}{2} + \frac{1}{2})^2 \le 1$. By Lemma~\ref{lemma:embed_d_gamma}, we
have $\ip{w_2/2}{x_s} = \ip{u_t/2}{v_t} \ge \gamma/2, \forall s \ge t$ and
$\ip{w_2/2}{x_s} = \ip{u_t/2}{v_s} \le - \gamma/2$ for all $s<t$. Similarly,
$\ip{w_1/2}{x_s} \le -\gamma/2$ for all $s \ge t$ and $\ip{w_1/2}{x_s} \ge
\gamma/2$ for all $s < t$. Thus, the examples are strongly separable with margin
$\gamma$.
\end{proof}
